Asymmetric relationships and their effects on coexistence

### 2023-09-18 paper/Paper.Rnw
cut main text length down below 5000

\noindent \textbf{Running head:} Asymmetric relationships and coexistence \\
\noindent \textbf{Key words:} asymmetric tail association, biodiversity, coexistence, competition, copula, modern coexistence theory, niche differentiation, storage effects \\
\noindent \textbf{Type of article:} Letter\\
\noindent \textbf{Words in Abstract:} 150; \textbf{Words in main text:} 4975; \textbf{Words in text boxes:} 0\\
\noindent \textbf{Number of references:} 39; \textbf{Number of figures:} 5; \textbf{Number of tables:} 1; \textbf{Number of text boxes:} 0\\
\noindent \textbf{Corresponding author:}
Daniel C. Reuman,
Department of Ecology and Evolutionary Biology,
University of Kansas,
2101 Constant Ave,
Lawrence, KS 66047;
Phone: 626-560-7084 \\
\noindent \textbf{Author emails:} pimsupaa@uoregon.edu, reuman@ku.edu\\
\noindent \textbf{Statement of Authorship:} The authors jointly carried out all parts of the study.\\
\noindent \textbf{Data accessibility statement:} All codes associated with this study have been released in a 
public repsitory.

@@

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists may ask, for instance, how local environment relates to 
population growth rate. To study relationships, ecologists 
very commonly use metrics such as correlation. But standard measures of the strength 
and direction of association between variables do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations are also often used (e.g., various forms of 
regression), but other common methods also reveal only a small portion 
of the information content of the relationship \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recent 
insights have been gained by studying how features of associations beyond standard measures 
impact ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. For instance, \cite{Ghosh_2020_Eco} argued that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting rescue effects. 
And \cite{Ghosh_2021} showed that ATAs can alter the stability of 
ecosystem functioning. 
Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs using 
plankton population density time series;
these examples are intended only to empirically illustrate what ATAs look like, as they
use data of a different type from what we focus on in this paper.
Many other examples of ATAs in nature have also been explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition have not been investigated. Doing so is our goal here.

%Intro to competition, coexistence, MCT, etc.
Competition and coexistence are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}. The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will 
instead drive the other extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems,
through the niche partitioning effects of environmental variation.
%moved the following two sentences from below
MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Letten2018}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is somewhat mathematically complex, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.
%
MCT breaks down coexistence into the contributions of various mechanisms, with names such as 
``storage effects'' \citep{Chesson_1994}. 
Storage effects, which are reviewed conceptually below and defined formally in Theory, refer to
a type of niche partitioning which allows species to effectively ``store" population growth 
benefits from favorable time periods in order to persist through unfavorable periods.

%Intro to storage effects and why they might be influenced by ATAs
We indicate why it is plausible that
ATAs may influence one mechanism of coexistence in particular, storage effects, starting by conceptually reviewing 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}. 
Storage effects allow competing species to fluctuate in abundance, without any going extinct,
by providing a relative benefit to whichever species is currently rare \citep{Chesson_2000}.
Storage effects hinge on covariation between the fluctuating quality of the environment
for a species ($E$) and the strength of competition ($C$) experienced by that species (called $EC$ covariance);
and on the assumption that no two species respond in the same way to environmental variation. Then, good environmental
conditions for a currently common species are paired with strong intraspecific competition, because of
positive $EC$ covariance and the commonness of the species. But good environmental conditions for 
a currently rare species are not as closely accompanied by strong competition because the rarity of
the species limits competitive influence. The resulting asymmetry is what provides the relative benefit to the
rare species. Storage effects can be quantified by measuring to what extent $EC$ covariance contributes to 
differences between growth rates of currently rare versus currently common species \citep{Chesson_1994,Ellner_2016}.  
ATAs may, in principle, influence storage effects simply because storage effects hinge on $EC$ covariation, 
and ATAs can be an aspect of $EC$ covariation. One scenario in which it is reasonable to imagine that
ATAs may be present in $EC$ covariation pertains to water limitation. Suppose two plants compete primarily for 
water, when water is scarce, but their niches become more differentiated when water is abundant.
Then the quality of the environment, $E$, as experienced by whichever plant is currently rare may only correspond closely to the 
competition, $C$, imposed on it by the currently abundant plant when water is scarce and $E$ is low.

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence, we carried out 
simple simulations using a two-species model, the lottery model. The simulations described in this
paragraph demonstrate that ATAs influence coexistence in some manner; the
nature of that influence is the subject of the rest of the paper.
The lottery model, which was introduced as a model of coral reef fish competing for a set of $N$ territories, 
is as follows.
Letting $N_i(t)$ denote the adult population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is an adult mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year. Note that $N = N_1(t)+N_2(t)$ 
is constant. 
We assume that the random variables $(B_1(t),B_2(t))$ are independent and identically 
distributed (iid) through time. In three distinct simulations (Fig. \ref{fig:pedagog1}f-h), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symmetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions were used in all 
simulations (see Methods
for details). Thus, the setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of species coexistence in these simulations (Fig. \ref{fig:pedagog1}f-h).
Notation for the paper is summarized in Table \ref{tab:notation}.
Typesetting of mathematics by ecological journals is sub-optimal,
so we appended a latex version of this main text to the end of the 
Supporting Information (SI); readers interested in the math should read that version.
%***DAN: Don't forget to do this

%Some intuition
We next give some initial intuition about how ATA effects on coexistence work, 
with more details in the Discussion.
It is known that storage effects can occur if a buffering mechanism reduces population 
losses in bad years so that
gains in good years can more than make up for them. 
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this can occur
both 1) when $E$ is moderately favorable but $C$ is very strong, and 2) when $E$ is 
unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this can occur 1) when $E$ is favorable and $C$ is not too strong,
and 2) when $E$ is moderately unfavorable but $C$ is very weak. 
It turns out that the potential for storage events to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (the cases 2) are not necessarily equal -- storage potential may differ
across the spectrum from 1 to 2.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with differences in the potential for storage across the spectrum from 1 to 2.
If $E$ and $C$ are largely unassociated in the parts of their distribution where
storage potential is high, ATA effects will be high. ATA effects quantify how temporal niche partitioning
varies across the $EC$ spectrum spanned by the cases 1-2 sketched above.

%Goal para
We previously demonstrated that ATAs are common in nature and can influence ecological phenomena
\citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}; and our arguments above
indicate that ATAs may influence coexistence. Thus, we here seek to quantify the consequences 
of ATAs for competition and coexistence, via the following specific goals. 
G1) We augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
G2) We explore, using the lottery model, the relative magnitudes 
of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence; and we determine whether species may 
sometimes coexist with ATAs but not without, or vice versa. This is to help 
illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
G3) We provide an empirical example of species coexistence which highlights the role ATAs can play and the
potential importance of ATAs.
The example is a laboratory diatom system. In the Discussion, we consider whether climate change may alter
ATAs of environmental variables and thereby influence competition.
Overall, our study presents a new mechanism of species coexistence
and a means of understanding its importance.

@@

We here pursue goal G1 from the Introduction: to augment MCT to show the role of ATAs for coexistence.

@@

We first develop our theory for a general two-species model.
Our notation parallels
that of \cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environmental factor $E_i(t)$ and as a decreasing function of 
competitive pressure on species $i$, $C_i(t)$, i.e., $r_i(E_i,C_i)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though this assumption is probably unnecessary, it simplifies the
presentation of new ideas.

@@

MCT and its computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that only one mechanism relates to ATAs: storage effects. 
So we begin by defining storage effects. 
The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between rare and common species mean growth rates, when $i$ is rare.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ is the mean ``growth when rare'' (GWR) of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ is the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (this is $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition have been rendered independent.  
A subscript 
$i \bs i$ always indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
a subscript $j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species. See \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ that we use; but one definition is $q_{ij}=1$, and to understand the new
ideas presented here it is sufficient to use that definition. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.

@@

We now specify how to decompose storage effects into
contributions due to ATAs and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between rare and common species mean growth rates for the original model and for a 
modified model for which $E$ and $C$
were rendered independent [see (\ref{eq:storage_effects_gen})]. 
Because components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship to the rare/common species growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the $EC$ relationship 
altered. Specifically, ATAs are removed -- see below for how. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the \emph{ATA effect on coexistence}. 
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
Note that brackets, $[\cdot]$, and parentheses, $(\cdot)$, are used distinctly
throughout the paper - they are not interchangeable.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (ATAs).
The symbol should correspondingly be pronounced ``partial sharp''.
See SI section \ref{SI-SIsect:techcomm} for technical comments on the notation. 
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are a term in the MCT decomposition \citep{Ellner_2019} -- see 
below for additional details.

>>>>

\noindent \textbf{Running head:} Asymmetric relationships and coexistence \\
\noindent \textbf{Key words:} asymmetric tail association, biodiversity, coexistence, competition, copula, modern coexistence theory, niche differentiation, storage effects \\
\noindent \textbf{Type of article:} Letter\\
\noindent \textbf{Words in Abstract:} 150; \textbf{Words in main text:} 4998; \textbf{Words in text boxes:} 0\\
\noindent \textbf{Number of references:} 39; \textbf{Number of figures:} 5; \textbf{Number of tables:} 1; \textbf{Number of text boxes:} 0\\
\noindent \textbf{Corresponding author:}
Daniel C. Reuman,
Department of Ecology and Evolutionary Biology,
University of Kansas,
2101 Constant Ave,
Lawrence, KS 66047;
Phone: 626-560-7084 \\
\noindent \textbf{Author emails:} pimsupaa@uoregon.edu, reuman@ku.edu\\
\noindent \textbf{Statement of Authorship:} The authors jointly carried out all parts of the study.\\
\noindent \textbf{Data accessibility statement:} All codes associated with this study have been released in a 
public repository.

@@

%Intro the paper by intro'ing ATAs
At the core of ecology is the study of relationships between organisms and their environments.
Ecologists ask, for instance, how local environment relates to 
population growth rate. To study relationships, ecologists 
commonly use metrics such as correlation. But standard measures of the strength 
and direction of association between variables do not tell the whole story 
\citep{Ghosh_2020_AER}.
To illustrate, the three pairs of variables in Fig. \ref{fig:pedagog1}a-c have the same 
Pearson correlation despite striking differences in association structure: Fig. \ref{fig:pedagog1}b shows a 
symmetric type of association, while Fig. \ref{fig:pedagog1}a,c exhibit what has been called 
``asymmetric tail association'' [ATA; \citealp{Ghosh_2020_AER}]. 
Metrics of association other than correlations are also used (e.g., various forms of 
regression), but other common methods also reveal only a small portion 
of the information in the relationship \citep{Nelsen2006}.
ATAs and related ideas had previously been seldom considered in ecology, but recently, 
insights were gained by studying how features of associations beyond standard measures 
impact ecological phenomena \citep{deValpine_2014, Anderson_2019, Popovic_2019, Ghosh_2020_AER, Ghosh_2020_Eco,
Ghosh_2020_ecolevol, Ghosh_2021}. \cite{Ghosh_2020_Eco} argued that extinction risks can become elevated 
when metapopulations are subject to ATAs, e.g., if environmental ``catastrophes'' are more spatially synchronous 
across component populations
than are years of good environment, limiting rescue effects. 
And \cite{Ghosh_2021} showed that ATAs alter the stability of 
ecosystem functioning. 
Fig. \ref{fig:pedagog1}d,e show contrasting examples of ATAs using 
plankton population density time series;
these examples are intended only to empirically illustrate what ATAs look like, as they
use data of a different type from the focus of this paper.
Many other examples of ATAs in nature have also been explored 
\citep{Ghosh_2020_AER,Ghosh_2020_ecolevol,Ghosh_2021}. To our knowledge, the implications of ATAs for 
competition have not been investigated. Doing so is our goal here.

%Intro to competition, coexistence, MCT, etc.
Competition and coexistence are topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}. The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will 
drive the other extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist in real ecosystems,
through the niche partitioning effects of environmental variation.
%moved the following two sentences from below
MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Caceres_1997, Descamps_2005, Letten2018}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.
%
MCT breaks down coexistence into the contributions of various mechanisms, with names such as 
``storage effects'' \citep{Chesson_1994}. 
Storage effects, which are reviewed conceptually below and defined formally in Theory, refer to
a type of niche partitioning which allows species to effectively ``store" population growth 
benefits from favorable periods in order to persist through unfavorable periods.

%Intro to storage effects and why they might be influenced by ATAs
We indicate why it is plausible that
ATAs may influence one mechanism of coexistence in particular, storage effects, starting by conceptually reviewing 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}. 
Storage effects allow competing species to fluctuate in abundance, without any going extinct,
by providing a relative benefit to whichever species is currently rare \citep{Chesson_2000}.
Storage effects hinge on covariation between the fluctuating quality of the environment
for a species ($E$) and the competition ($C$) experienced by that species (called $EC$ covariance);
and on the assumption that no two species respond identically to environmental variation. Then, good environmental
conditions for a currently common species are paired with strong intraspecific competition, because of
positive $EC$ covariance and the commonness of the species. But good environmental conditions for 
a currently rare species are not as closely accompanied by strong competition because the rarity of
the species limits competitive influence. The resulting asymmetry is what provides a benefit to the
rare species. Storage effects can be quantified by measuring how much $EC$ covariance contributes to 
differences between growth rates of currently rare versus currently common species \citep{Chesson_1994,Ellner_2016}.  
ATAs may, in principle, influence storage effects simply because storage effects hinge on $EC$ covariation, 
and ATAs can be an aspect of $EC$ covariation. One scenario in which it is reasonable to imagine that
ATAs may be present in $EC$ covariation pertains to water limitation. Suppose two plants compete primarily for 
water, when water is scarce, but their niches become differentiated when water is abundant.
Then the quality of the environment, $E$, as experienced by whichever plant is currently rare, may only correspond closely to the 
competition, $C$, imposed on it by the currently abundant plant when water is scarce and $E$ is low.

%Lottery model and results based on it
To further demonstrate the plausibility of ATAs altering coexistence, we carried out 
simulations using a two-species model, the lottery model. The simulations described in this
paragraph demonstrate that ATAs influence coexistence in some manner; the
nature of that influence is the subject of the rest of the paper.
The lottery model, which was introduced as a model of coral reef fish competing for a set of $N$ territories, 
is as follows.
Letting $N_i(t)$ denote the adult population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is an adult mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that adults die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the species that year. Note that $N = N_1(t)+N_2(t)$ 
is constant. 
We assume that the random variables $(B_1(t),B_2(t))$ are independent and identically 
distributed (iid) through time. In three distinct simulations (Fig. \ref{fig:pedagog1}f-h), 
$B_1$ and $B_2$ were, respectively, 
left- (Fig. \ref{fig:pedagog1}a), right- (Fig. \ref{fig:pedagog1}c) and symmetrically (Fig. \ref{fig:pedagog1}b)
tail associated, while the same marginal distributions were used in all 
simulations (see Methods
for details). Thus, the setup was identical in the three simulations except for different ATAs. 
ATAs substantially influenced aspects of coexistence in these simulations (Fig. \ref{fig:pedagog1}f-h).
Notation for the paper is summarized in Table \ref{tab:notation}.
Typesetting of mathematics by ecological journals is sub-optimal,
so, for mathematical readers, we appended a latex version of this main text to the end of the Supporting Information (SI).
%***DAN: Don't forget to do this

%Some intuition
We next give some intuition about how ATA effects on coexistence work, 
with details in the Discussion.
It is known that storage effects can occur if a buffering mechanism reduces population 
losses in bad years so that
gains in good years can more than make up for them. 
Bad years happen when the negative effects of $C$ outweigh any 
positive effects of $E$; but this occurs
both 1) when $E$ is moderately favorable but $C$ is very strong, and 2) when $E$ is 
unfavorable and $C$ is not too weak.
Conversely, good years happen when positive effects of $E$ outweigh negative effects of $C$;
but this occurs 1) when $E$ is favorable and $C$ is not too strong,
and 2) when $E$ is moderately unfavorable but $C$ is very weak. 
It turns out that the potential for storage events to occur when 
$E$ and $C$ are both higher than average (the cases 1, above); versus the potential  
when $E$ and $C$ are
both lower than average (cases 2) are not necessarily equal -- storage potential may differ
across the spectrum from 1 to 2.
ATA effects quantify the extent to which asymmetries of association between $E$ and 
$C$ are aligned with differences in the potential for storage across the spectrum from 1 to 2.
If $E$ and $C$ are largely unassociated in the parts of their distribution where
storage potential is high, ATA effects will be high. ATA effects quantify how temporal niche partitioning
varies across the $EC$ spectrum spanned by the cases 1-2 above.

%Goal para
We previously demonstrated that ATAs are common in nature and can influence ecological phenomena
\citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}; and our arguments above
indicate that ATAs may influence coexistence. Thus, we here seek to quantify the consequences 
of ATAs for competition and coexistence, via the following specific goals. 
G1) We augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
G2) We explore, using the lottery model, the relative magnitudes 
of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence; and we determine whether species may 
sometimes coexist with ATAs but not without, or vice versa. This is to help 
illuminate whether ATA contributions to coexistence 
are negligible or meaningfully large. 
G3) We provide an empirical example of species coexistence which highlights the role ATAs can play and their
potential importance.
The example is a laboratory diatom system. In the Discussion, we consider whether climate change may alter
ATAs of environmental variables and thereby influence competition.
Overall, our study presents a new mechanism of species coexistence
and a means of understanding its importance.

@@

We here pursue goal G1: to augment MCT to show the role of ATAs for coexistence.

@@

We first develop our theory for a general two-species model.
Our notation parallels
\cite{Ellner_2016,Ellner_2019}. It is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environmental factor $E_i(t)$ and as a decreasing function of 
competitive pressure on species $i$, $C_i(t)$, i.e., $r_i(E_i,C_i)$. We also assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though this assumption is probably unnecessary, it simplifies the
presentation of new ideas.

@@

MCT and its computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that only one mechanism relates to ATAs: storage effects. 
So we begin by defining storage effects. 
The storage effect for species $i$ is the contribution of covariation between $E$ and 
$C$ to a difference between rare and common species mean growth rates, when $i$ is rare.
Specifically, storage effects for species $i$ are 
\begin{equation}
\Delta_i^{(EC)} = (\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i}),\label{eq:storage_effects_gen}
\end{equation}
where: $\olr_{i \bs i}$ is the mean ``growth when rare'' (GWR) of species $i$, that is, the mean growth rate of 
species $i$ when it is at negligible abundance and species $j$ (the other species) is at stochastic steady state;
$\olr_{j \bs i}$ is the mean steady-state 
growth rate of species $j$ when species $i$ is absent or at negligible abundance (equals $0$, \emph{a priori});
and $\olrs_{i \bs i}$ and $\olrs_{j \bs i}$ are analogous quantities, but for an alternative modelling scenario, identical except that
influences of environment and competition have been rendered independent.  
A subscript 
$i \bs i$ indicates a quantity computed for species $i$ when that species is at negligible abundance, and 
$j \bs i$ indicates a quantity computed for $j$ when $i$ is absent.
The \emph{scaling factor}
$q_{ij}$ measures the relative sensitivities to competition of the two species. See \cite{Chesson_1994},
\cite{Ellner_2016} and SI section \ref{SI-SIsect:scaling_factors} for the two
alternative definitions of $q_{ij}$ we use; but one definition is $q_{ij}=1$, and to understand the 
ideas presented here it is sufficient to use that definition. 
Since it was assumed that $r_i=r_i(E_i,C_i)$, we can write $\olr_{i \bs i}=\E[r_i(E_i,C_{i \bs i})]$, where
$\E(\cdot)$ denotes expected value just as does an overbar, but we use $\E$ for longer expressions.
Also, $\olr_{j \bs i}=\E[r_j(E_j,C_{j \bs i})]$.
An expression for $\olrs_{i \bs i}$ is obtained from $\olr_{i \bs i}$ by replacing the bivariate random 
variable $(E_i,C_{i \bs i})$ by a random variable with the same marginal distributions, but independent components, 
$(E_i^\shp,C_{i \bs i}^\shp)$, so $\olrs_{i \bs i}=\E[r_i(E_i^\shp,C_{i \bs i}^\shp)]$.
Likewise, $\olrs_{j \bs i}=\E[r_j(E_j^\shp,C_{j \bs i}^\shp)]$.

@@

We now specify how to decompose storage effects into
contributions due to ATAs and 
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between rare and common species mean growth rates for the original model and for a 
modified model for which $E$ and $C$
were rendered independent [see (\ref{eq:storage_effects_gen})]. 
Because components were 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship to the rare/common species growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the $EC$ relationship 
altered. Specifically, ATAs are removed -- see below for how. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[EC]} + \Delta_i^{[E \nat C]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[EC]}$ is the \emph{ATA effect on coexistence}. 
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to $EC$ 
correlation \emph{per se}, remaining after having removed the effects of ATAs.
Note that brackets, $[\cdot]$, and parentheses, $(\cdot)$, are used distinctly
throughout the paper - they are not interchangeable.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (ATAs).
The symbol should be pronounced ``partial sharp''.
See SI section \ref{SI-SIsect:techcomm} for technical comments on the notation. 
The brackets in $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are a term in the MCT decomposition \citep{Ellner_2019} -- see 
below for details.

How are asymmetries of tail association removed while leaving $E$ and $C$ marginal distributions and overall 
correlation unchanged? Fig. \ref{fig:pedagog2} provides an illustration of the steps we describe here. 
We assume, for simplicity, that the cumulative distribution functions (CDFs) of $E_i$ and 
$C_{i \bs i}$, which we denote $F_{E_i}$ and $F_{C_{i \bs i}}$, are strictly monotonic and continuous. 
Letting $\varphi$ be the CDF of a standard normal distribution, we consider the random variables
$\varphi^{-1} \circ F_{E_i} (E_i)$ and $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})$, which are standard normally
distributed, although they are not necessarily jointly bivariate normal (Fig. \ref{fig:pedagog2}a,b).  
Let $\rho_i$ be their covariance. We then define a bivariate normal random variable $(e_i,c_{i \bs i})$ such that
$\E(e_i)=\E(c_{i \bs i})=0$ and $\var(e_i)=\var(c_{i \bs i})=1$, and we denote 
$\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$ (Fig. \ref{fig:pedagog2}c). We define 
$E_i^\nat = F_{E_i}^{-1} \circ \varphi(e_i)$
and $C_{i \bs i}^\nat=F_{C_{i \bs i}}^{-1} \circ \varphi(c_{i \bs i})$ (Fig. \ref{fig:pedagog2}d). 
The value of $\ddot{\rho}_i$ is determined in one of 
four ways: either by setting $\ddot{\rho}_i = \rho_i$; or so that the Pearson, Spearman or Kendall correlation of 
$E_i^\nat$ and $C_{i \bs i}^\nat$ matches that of $E_i$ and $C_{i \bs i}$. The choice of bivariate normal 
$(e_i,c_{i \bs i})$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are symmetrically tail associated.
For those familiar with copula statistics (see, e.g., \citealp{Genest_2007} and \citealp{Ghosh_2020_AER}), 
we replaced the copula of $(E_i,C_{i \bs i})$ with a normal copula to form $(E_i^\nat,C_{i \bs i}^\nat)$.
The choice of $\ddot{\rho}_i$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are correlated to the same extent as 
$E_i$ and $C_{i \bs i}$, according to whichever definition of correlation is preferable. Thus $E_i^\nat$ and 
$C_{i \bs i}^\nat$ have had their tail association rendered symmetric while retaining their correlation \emph{per se}. 
The random variable $(E_j^\nat, C_{j \bs i}^\nat)$ is created similarly.  
For comparison, Fig. \ref{fig:pedagog2} also shows, for the example explored by that figure, 
the distribution $(E_i^\shp,C_{i \bs i}^\shp)$
defined above (following \cite{Ellner_2016}).
